---
title: "Practical Machine Learning - Final Project"
author: "Cristian HT"
date: "16/12/2020"
output: 
        html_document:
         keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

This is the final report for Coursera’s Practical Machine Learning course, as part of the Data Science Specialization track offered by John Hopkins. 

In this project, we will use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants to predict the manner in which they did the exercise. This is the “classe” variable in the training set. We train 4 models: **Decision Tree**, **Random Forest**, **Gradient Boosted Trees**, **Support Vector Machine** using k-folds cross validation on the training set. We then predict using a validation set randomly selected from the training csv data to obtain the **accuracy** and **out of sample error rate**. Based on those numbers, we decide on the best model, and use it to predict 20 cases using the test csv set. 


# Data

The training data for this project are available here:

**https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv**

The test data are available here:

**https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv**


# Loading Data and Libraries

```{r results='hide', message=FALSE}
library(lattice)
library(ggplot2)
library(caret)
library(kernlab)
library(rattle)
library(corrplot)
set.seed(1234)
```

## Download Data

```{r}
URL_training="https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
URL_testing="https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
download.file(URL_training, "pml-training.csv")
download.file(URL_training,"pml-testing.csv")
```

## Read Data
```{r}
traincsv <- read.csv("pml-training.csv")
testcsv <- read.csv("pml-testing.csv")

dim(traincsv)
dim(testcsv)
```

We see that there are 160 variables and 19622 observations in the training set, while 20 for the test set.


# Cleaning the Data

Removing unnecessary variables. Starting with N/A variables.
```{r}
traincsv <- traincsv[,colMeans(is.na(traincsv)) < .9] #removing mostly na columns
traincsv <- traincsv[,-c(1:7)] #removing metadata which is irrelevant to the outcome
```

Removing near zero variance variables.
```{r nzv}
nvz <- nearZeroVar(traincsv)
traincsv <- traincsv[,-nvz]
dim(traincsv)
```

Now that we have finished removing the unnecessary variables, we can now split the training set into a **validation** and sub **training** set. The testing set "testcsv" will be left alone, and used for the final quiz test cases. 
```{r}
inTrain <- createDataPartition(y=traincsv$classe, p=0.7, list=F)
train <- traincsv[inTrain,]
valid <- traincsv[-inTrain,]
```


# Creating and Testing the Models
Here we will test a few popular models including: **Decision Trees**, **Random Forest**, **Gradient Boosted Trees**, and **SVM**. This is probably more than we will need to test, but just out of curiosity and good practice we will run them for comparison.

Set up control for training to use 3-fold cross validation. 
```{r}
control <- trainControl(method="cv", number=3, verboseIter=F)
```

## Decision Tree

**Model:** 

```{r, cache=TRUE}
mod_trees <- train(classe~., data=train, method="rpart", trControl = control, tuneLength = 5)
fancyRpartPlot(mod_trees$finalModel)
```

**Prediction:**

```{r}
pred_trees <- predict(mod_trees, valid)
cmtrees <- confusionMatrix(pred_trees, factor(valid$classe))
cmtrees
```

## Random Forest

```{r, cache=TRUE}
mod_rf <- train(classe~., data=train, method="rf", trControl = control, tuneLength = 5)

pred_rf <- predict(mod_rf, valid)
cmrf <- confusionMatrix(pred_rf, factor(valid$classe))
cmrf
```

## Gradient Boosted Trees

```{r, cache=TRUE}
mod_gbm <- train(classe~., data=train, method="gbm", trControl = control, tuneLength = 5, verbose = F)

pred_gbm <- predict(mod_gbm, valid)
cmgbm <- confusionMatrix(pred_gbm, factor(valid$classe))
cmgbm
```

## Support Vector Machine

```{r, cache=TRUE}
mod_svm <- train(classe~., data=train, method="svmLinear", trControl = control, tuneLength = 5, verbose = F)

pred_svm <- predict(mod_svm, valid)
cmsvm <- confusionMatrix(pred_svm, factor(valid$classe))
cmsvm
```

## Results (Accuracy & Out of Sample Error)

```{r, echo=FALSE}
models <- c("Tree", "RF", "GBM", "SVM")
accuracy <- round(c( cmtrees$overall[1], cmrf$overall[1], cmgbm$overall[1], cmsvm$overall[1]),3) #accuracy
oos_error <- 1 - accuracy #out of sample error

data.frame(accuracy = accuracy, oos_error = oos_error, row.names = models)
```

**The best model is the Random Forest model, with `r cmrf$overall[1]` accuracy and `r 1-cmrf$overall[1]` out of sample error rate. We find that to be a sufficient enough model to use for our test sets. ** 


# Predictions on Test Set

Running our test set to predict the classe (5 levels) outcome for 20 cases with the **Random Forest** model.
```{r}
pred <- predict(mod_rf, testcsv)
print(pred)
```


# Appendix

**Plotting the models**

```{r}
plot(mod_trees)
plot(mod_rf)
plot(mod_gbm)
```
